---
title: "MLR Analysis w/ Life Satisfaction" 
author: "Ruoyun Li"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE,warning=FALSE)
options(tinytex.verbose = TRUE)
```

## Data Ingestion & Transformation
```{r}
library(tidyverse)
#library(finalfit)

#ppdata = read.csv("surveydata.csv")
load("/Users/ruoyunrosa/Desktop/STAT 4996/SurveyDF.rda")
data <- da38964.0001


# extract numbers inside parentheses
extract_number <- function(x) {
  gsub("\\((\\d+)\\) .*", "\\1", x)
}
# apply the function to column 4-372
data[, 4:372] <- lapply(data[, 4:372], extract_number)

# convert character into integer
data[, 4:372] <- lapply(data[, 4:372], function(x) as.integer(x))
data[, 4:372] <- as.data.frame(data[, 4:372])

# select variables of interest
data = select(data, RACEREC, HOUSING_2, DRUG_DIS, RUCC, ALC_DIS, SEX, PHYS_HEALTH, EDUC, MILITARY, SLEEPHRS, MARSTAT, AGE, HHINC, MENT_HEALTH, EMP_1, EMP_2, STUDENT, COV_EXTUI, COV_STIMULUS, COV_RENTMORT, COV_OTHERPAYMENT, COV_MEDBILLS, COV_MONEYFOOD, COV_CREDITDEBT, SAT1, SAT2, SAT3, SAT4, SAT5, DOWN, WORRY, INTEREST, NERVOUS, FINAL_WGT)


# data type and level validation
data = data %>%
  mutate_at(vars(RACEREC, HOUSING_2, DRUG_DIS, RUCC, ALC_DIS, SEX, PHYS_HEALTH, EDUC, MILITARY, SLEEPHRS, MARSTAT, AGE, HHINC, MENT_HEALTH, EMP_1, EMP_2, STUDENT, COV_EXTUI, COV_STIMULUS, COV_RENTMORT, COV_OTHERPAYMENT, COV_MEDBILLS, COV_MONEYFOOD, COV_CREDITDEBT), as.factor) %>%
  mutate_at(vars(DOWN, WORRY, INTEREST, NERVOUS, SAT1, SAT2, SAT3, SAT4, SAT5, AGE, FINAL_WGT), as.numeric)

data$SEX = factor(data$SEX, levels = c(1, 2, 3), labels = c("Male", "Female", "Other"))
data$MARSTAT = factor(data$MARSTAT, levels = c(1, 2, 3,4,5,6,99), labels = c("Married", "Divorced", "Separated", "Widowed", "Single", "Unmarried couple", "Refusal"))


cols_to_recode <- c("COV_RENTMORT", "COV_STIMULUS", "COV_OTHERPAYMENT", 
                    "COV_MEDBILLS", "COV_MONEYFOOD", "COV_CREDITDEBT")

# Recode values for all specified columns
data <- data %>%
  mutate(across(all_of(cols_to_recode), ~ recode(.,
                                                 `1` = "Yes",
                         `2` = "No",
                         `99` = "Refusal")))


data$RACEREC = factor(data$RACEREC, levels = c(1,2,3,4,5,6,98))
data$DRUG_DIS = factor(data$DRUG_DIS, levels = c(1,2,98,99), labels = c("Yes", "No", "Don't Know", "Refusal"))
data$RUCC = factor(data$RUCC, levels = c(1,2,3,4,5,6,7,8,9))
data$ALC_DIS = factor(data$ALC_DIS, levels = c(1,2,98,99), labels = c("Yes", "No", "Don't Know", "Refusal"))
data$PHYS_HEALTH = factor(data$PHYS_HEALTH, levels = c(1,2,3,4,5), labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))
data$EDUC = factor(data$EDUC, levels = c(1,2,3,4,5,6))
data$MILITARY = factor(data$MILITARY, levels = c(1,2,3,4,99), labels = c("Never", "Current", "Past", "Training", "Refusal"))
data$HHINC = factor(data$HHINC, levels = c(1,2,3,4,5,6,7,8,9,10,11,12,98))

data <- data %>% 
  mutate(across(SAT1:SAT5, ~ case_when(
    . %in% 1:5 ~ as.numeric(.),
    . %in% c(6, 7, 98, 99) ~ NA_real_,
    TRUE ~ NA_real_
  )))
data <- data %>% 
  mutate(across(c(WORRY, DOWN, INTEREST, NERVOUS), ~ case_when(
    . %in% 1:4 ~ as.numeric(.),
    . %in% c(5,6, 98, 99) ~ NA_real_,
    TRUE ~ NA_real_
  )))
# turn the worry variables scale into PHQ4 corresponding scale
data <- data %>%
  mutate(across(c(DOWN, WORRY, INTEREST, NERVOUS), ~ case_when(
    . == 1 ~ 0,
    . == 2 ~ 1,
    . == 3 ~ 2,
    . == 4 ~ 3,
    TRUE ~ NA_real_ # Keep any unexpected values as NA
  )))
###

# SAT variable creation
data = data %>% mutate(SAT_total = ifelse(rowSums(select(., SAT1:SAT5) == 99, na.rm = TRUE) > 0, NA, rowSums(select(., SAT1:SAT5), na.rm = TRUE)))

# PHQ variable creation
data = data %>% mutate(across(c(DOWN, WORRY, INTEREST, NERVOUS), ~ ifelse(. %in% c(98,99), NA, .)))
data = data %>% mutate(PHQ = rowSums(select(., DOWN, WORRY, INTEREST, NERVOUS), na.rm = TRUE))

# EMPLOYMENT variable creation
data = data %>% mutate(EMP = factor(paste(EMP_1, EMP_2, STUDENT, sep = "_"), levels = c("0_0_0", "0_0_1", "0_1_0", "0_1_1", "1_0_0", "1_0_1", "1_1_1"), labels = c("None", "Student", "EmployedMultiple", "EmployedMultiple and Student", "EmployedSingle", "EmployedSingle and Student", "All")))

# COVID rentmort + other variable creation
data = data %>% mutate(COV_MISSPAY = if_else(COV_RENTMORT == "Yes" | COV_OTHERPAYMENT == "Yes", "Yes", "No"))
data$COV_MISSPAY <- factor(data$COV_MISSPAY)

data$SLEEPHRS<-as.numeric((data$SLEEPHRS))
data$FINAL_WGT<-as.numeric((data$FINAL_WGT))
```


```{r}
### Replace refusal, 98, 99, 100 with NA
data[data == "Refusal" | data == 98 | data == 99 | data == 100 | data == 96] <- NA
colSums(is.na(data))

data <- na.omit(data)
```


```{r}
# Remove unnecessary columns
data$SAT1<-NULL
data$SAT2<-NULL
data$SAT3<-NULL
data$SAT4<-NULL
data$SAT5<-NULL
data$DOWN<-NULL
data$NERVOUS<-NULL
data$INTEREST<-NULL
data$WORRY<-NULL

data$COV_RENTMORT<-NULL
data$COV_OTHERPAYMENT<-NULL
data$STUDENT<-NULL
data$EMP_1<-NULL
data$EMP_2<-NULL
```


## MLR w/ Life Satisfaction 

Build demographic model → add financial variables and run partial f test
- motivation: demographic served as control. to see whether financial variables improve the model. 
- forward selection 
- variable selection: expert, literature review, 
Or demographic → forward backward → add all financial and run partial f test
Significant → run forward backward with fixed demographic variables and check financial variables


```{r}
#library(GGally)
library(dplyr)

dataSat<-data
dataSat$PHQ<-NULL

data.no.wt <- dataSat %>% select(-FINAL_WGT)

hist(data.no.wt$SAT_total)
```


## Checking MLR assumptions 

Five main assumptions underlying multiple regression models must be satisfied: (1) linearity, (2) homoscedasticity, (3) independence of errors, (4) normality, and (5) independence of independent variables.

```{r}
result<-lm(SAT_total~., data=data.no.wt, weights = data$FINAL_WGT)
par(mfrow=c(2,2))
plot(result)
```

The first plot (top left) is the residual plot, with residuals on the y-axis and fitted values on the x-axis. The residual plot can be used to address assumptions 1 and 2. A red line is overlayed to represent the average value of the residuals for differing values along the x-axis. This line should be along the x-axis without any apparent curvature to indicate the form of our model is reasonable. This is generally the case for this regression. This addresses point 1. This is the most important assumption to check. For assumption 2, we want to see the vertical spread of the residuals to be fairly constant, which we have.

The second plot (top right) is the normal probability plot (also called a QQ plot), and addresses assumption 4. If the residuals are normal, the residuals should fall along the 45 degree line. The regression model is fairly robust to this assumption though; the normality assumption is the least crucial of the four.

The third plot (bottom left) is a plot of the square root of the standardized residuals against the fitted values (scale-location). This plot should be used to assess assumption 2, the constant variance assumption. If the variance is constant, the red line should be horizontal and the vertical spread of plot should be constant. This is generally the case for this regression.

The last plot (bottom right) is a plot to identify influential outliers. Data points that lie in the contour lines with large Cook’s distance are influential. 

Assumption 3, the independence of errors, is assessed by understanding the nature of the data. A by-product of this assumption is the the observed values of the response variable are independent. 

### ACF

```{r}
# autocorrelation (ACF) plot for assumption 3
acf(result$res, main="ACF Plot")
```
The correlations should be insignificant for all lags greater than 0. This is the case for this regression. The assumption 3 is met. 

We don’t see evidence of assumptions being violated. From the residual plot, we see an even scatter of residuals across the horizontal axis, as well as fairly constant variance.

Given that we have a number of predictors, we suspect that multicollinearity may be present.


### VIFs for multicollinearity
```{r}
library(car)
vif(result)
```
VIFs above 5 indicate a moderate degree of multicollinearity, while VIFs above 10 indicate a strong degree of multicollinearity.


### ANOVA F test and individual t tests
F test is significant. 
```{r}
summary(result)
```

## Model building

### Start with demographic var
Significant F test
```{r}
demo_vars <- c("SEX", "EDUC", "RACEREC", "EMP", "MARSTAT", "HOUSING_2", 
               "RUCC", "PHYS_HEALTH", "HHINC", "MENT_HEALTH","AGE", 
               "MILITARY","SLEEPHRS")
cat_demo <- c("SEX", "EDUC", "RACEREC", "EMP", "MARSTAT", "HOUSING_2", 
               "RUCC", "PHYS_HEALTH", "MILITARY", "HHINC")
# Convert categorical variables to factors
data.no.wt[cat_demo] <- lapply(data.no.wt[cat_demo], as.factor)

# Create a model with only demographic variables
demo_model <- lm(SAT_total ~ ., data = data.no.wt[, c("SAT_total", demo_vars)], weights = data$FINAL_WGT)
summary(demo_model)
```

### Model comparison

### Forward Selection on demographic variables
```{r}
# Backward selection
##intercept only model
regnull <- lm(SAT_total~1, data=data.no.wt, weights = data$FINAL_WGT)
##model with all predictors
regfull <- lm(SAT_total ~ SEX + EDUC + RACEREC + EMP + MARSTAT + HOUSING_2 + 
                 SLEEPHRS + RUCC + PHYS_HEALTH + MILITARY + HHINC + MENT_HEALTH + AGE, 
                 data = data.no.wt, weights = data$FINAL_WGT)

backward_model<-step(regfull, scope=list(lower=regnull, upper=regfull), direction="backward")
summary(backward_model)
```


```{r}

# forward selection
forward_model<-step(regnull, scope=list(lower=regnull, upper=regfull), direction="forward")
summary(forward_model)
```
Forward and backward selection select the same sets of variables. We will continue with the model: SAT_total ~ SEX + EMP + MARSTAT + HOUSING_2 + PHYS_HEALTH + 
    HHINC + MENT_HEALTH + AGE



### Add Financial Variables

```{r}
fin_vars <- c("COV_EXTUI", "COV_STIMULUS", "COV_MEDBILLS", "MONEYFOOD", "CREDITDEBIT") 
demoFin_result <- lm(SAT_total~SEX + EMP + MARSTAT + HOUSING_2 + PHYS_HEALTH + 
    HHINC + MENT_HEALTH + AGE+ COV_EXTUI+COV_STIMULUS+COV_MEDBILLS+ COV_MONEYFOOD+COV_CREDITDEBT, data = data.no.wt, weights = data$FINAL_WGT)

summary(demoFin_result)
```

### Partial F test for financial reduced model
drop COV_STIMULUS, COV_EXTUI and  COV_MEDBILLS
```{r}
reducedFindemo <- lm(SAT_total~MENT_HEALTH + PHYS_HEALTH + MARSTAT + 
    HOUSING_2 + SEX + AGE + HHINC + SLEEPHRS + EDUC + EMP + COV_MONEYFOOD+COV_CREDITDEBT+COV_STIMULUS, data = data.no.wt, weights = data$FINAL_WGT)

anova(reducedFindemo, demoFin_result)
```
The p-value is greater than 0.05. So we fail to reject the null hypothesis, so there is little evidence of supporting the full model. We go with the reduced model over the full model.

### Partial F test to compare demo model and Findemo model
We reject the null hypothesis, so there is evidence of supporting the full model. We go with the model includes both demographic and financial variables. 

```{r}
anova(forward_model, reducedFindemo)
```

```{r}
summary(reducedFindemo)
```


### Check assumptions of our final model

```{r}
par(mfrow=c(2,2))
plot(reducedFindemo)
```

### Consider Interactions

The following variables have significant coefficients with p-values less than 0.05:

MENT_HEALTH categories (MENT_HEALTH2, MENT_HEALTH3, etc.)
PHYS_HEALTH categories (PHYS_HEALTHVery Good, PHYS_HEALTHGood, etc.)
MARSTAT categories (MARSTATDivorced, MARSTATSingle, etc.)
HOUSING_2 categories (HOUSING_22, HOUSING_23, etc.)
SEX (Female, Other)
AGE
SLEEPHRS
HHINC categories (HHINC8, HHINC9, etc.)
COV_MONEYFOOD (No)
COV_CREDITDEBT (No)

Mental Health × Physical Health: It's plausible that someone with worse mental health may experience a more significant drop in SAT scores if they also have poor physical health.

Income × Marital Status: The effect of household income on SAT scores may depend on the marital status of the individual. People with higher income may have different needs and responses depending on whether they are married or not.

Age × Sleep Hours: The relationship between age and SAT scores could depend on how many hours of sleep the individual gets. Younger individuals may have a different response to sleep hours compared to older individuals.

### One Hot Encoding
```{r}
data.no.wt.encoded = cbind(
  data.no.wt,
  model.matrix(~ MENT_HEALTH + PHYS_HEALTH + HHINC + MARSTAT + AGE + SLEEPHRS - 1, data = data.no.wt)
)

# CREATE INTERACTION TERMS
MENT_HEALTH_VARS <- c("MENT_HEALTH1", "MENT_HEALTH2", "MENT_HEALTH3", "MENT_HEALTH4", "MENT_HEALTH5")
PHYS_HEALTH_VARS <- c("PHYS_HEALTHVery Good", "PHYS_HEALTHGood", "PHYS_HEALTHFair", "PHYS_HEALTHPoor")
HHINC_VARS <- c("HHINC2", "HHINC3", "HHINC4", "HHINC5", "HHINC6", "HHINC7", "HHINC8", "HHINC9", "HHINC10", "HHINC11", "HHINC12")
MARSTAT_VARS <- c("MARSTATDivorced", "MARSTATSeparated", "MARSTATWidowed", "MARSTATSingle", "MARSTATUnmarried couple")
AGE_VAR <- "AGE"
SLEEP_HOURS_VAR <- "SLEEPHRS"

# Create interaction terms for mental health × physical health
MENTAL_HEALTH_PHYSICAL_INTERACTIONS <- paste(MENT_HEALTH_VARS, PHYS_HEALTH_VARS, sep="*")

# Create interaction terms for household income × marital status
INCOME_MARITAL_INTERACTIONS <- paste(HHINC_VARS, MARSTAT_VARS, sep="*")

# Create interaction term for age × sleep hours
AGE_SLEEP_INTERACTION <- paste(AGE_VAR, SLEEP_HOURS_VAR, sep="*")

# Combine all the interaction terms into a single list
INTERACTION_TERMS <- c(MENTAL_HEALTH_PHYSICAL_INTERACTIONS, INCOME_MARITAL_INTERACTIONS, AGE_SLEEP_INTERACTION)

# Build the formula for the model dynamically
FORMULA <- paste("SAT_total~MENT_HEALTH + PHYS_HEALTH + MARSTAT + HOUSING_2 + SEX + AGE + HHINC + SLEEPHRS + EDUC + EMP + COV_MONEYFOOD + COV_CREDITDEBT + COV_STIMULUS + ", 
                 paste(INTERACTION_TERMS, collapse = " + "), sep=" ")

print(FORMULA)

add_backticks <- function(formula_string) {
  # Regular expression to identify variable names with spaces
  modified_formula <- gsub("([A-Za-z0-9_]+[[:space:]][A-Za-z0-9_]+)", "`\\1`", formula_string)
  
  return(modified_formula)
}

modified_formula <- add_backticks(FORMULA)
print(modified_formula)
```

### reducedFindemo with interactions
```{r}
reducedFindemoINT <- lm(as.formula(modified_formula), data = data.no.wt.encoded, weights = data$FINAL_WGT)
anova(reducedFindemo, reducedFindemoINT)
```
We have a p-value greater than .05 so we fail to reject the null hypothesis of both models having similar model fit, there is no evidence to support adding interaction terms significantly improves the model.
```{r}
summary(reducedFindemoINT)
```

